Model-order reduction of non-linear time-dependent neutronics via POD-Galerkin projection and DEIM

### 2022-03-30 paper_/main.tex
added annals version

% interactnlmsample.tex
% v1.05 - August 2017

\documentclass[]{interact}

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage[caption=false]{subfig}% Support for small, `sub' figures and tables
%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
%\usepackage[doublespacing]{setspace}% To produce a `double spaced' document if required
%\setlength\parindent{24pt}% To increase paragraph indentation when line spacing is doubled

\usepackage{algorithm,algorithmic}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{caption}
\usepackage{threeparttable}
\usepackage{tikz}
\usepackage{multirow}
\usepackage{booktabs,tabularx}
%\usepackage{subcaption}
\DeclareMathOperator{\argmaxH}{argmax} 
\DeclareMathOperator*{\argmin}{argmin}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\newcommand\abs[1]{\left|#1\right|}

\usepackage[numbers,sort&compress]{natbib}% Citation support using natbib.sty
\bibpunct[, ]{[}{]}{,}{n}{,}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty
\makeatletter% @ becomes a letter
\def\NAT@def@citea{\def\@citea{\NAT@separator}}% Suppress spaces between citations using natbib.sty
\makeatother% @ becomes a symbol again

\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}
\bibliographystyle{plain}

\begin{document}

\articletype{Journal of Nuclear Science and Technology}% Specify the article type or omit as appropriate

\title{Model-order reduction of non-linear time dependent neutronics via POD-Galerkin projection and DEIM}
\author{
\name{Rabab Elzohery\thanks{CONTACT A.~N. Author. Email: rababelzohery@ksu.edu} and Jeremy Roberts}
\affil{Department of Mechanical and Nuclear Engineering,
	Kansas State University \\
	3002 Rathbone Hall, Manhattan, KS 66506, US \\ rababelzohery@ksu.edu, jaroberts@ksu.edu}
}

\maketitle
\begin{abstract}
Presented here is an application of model-order reduction to non-linear neutronic transients.
% where reduced-order models (ROMs) are generally sought to provide a rapid, yet accurate, approximation of a particular simulation response.
The effort described extends previous work\cite{elzohery2021modeling} aimed to develop and assess a ROM framework that can be used for different purposes, such as design optimization and propagating input parameter uncertainties like nuclear data.
Here, the target applications here is non-linear.
The ROM was built with the use of the POD-Galerkin projection method and the matrix version of the Discrete Empirical Interpolation Method (MDEIM) for improved efficiency.
The MDEIM is used to approximate the projection of the problem operator that otherwise must be computed at each non-linear iteration within each time step, which increases the ROM cost.
By using the MDEIM, an offline-online ROM decomposition was obtained, which is favored from a computational perspective since the expensive projection operation can take place only once in the offline stage.
The online stage, which involves solving a linear system  at each time step, depends only on the reduced dimension and is therefore cheap, making it possible to increase the solution time resolution, and accordingly accuracy, without introducing significant cost.
By applying the method to the LRA benchmark, speed ups from 6 to 40 were achieved depending on the full-order model (FOM) spatial fidelity.
The maximum relative error in the assemblies power was in the order of $1\times10^{-5}$\%.
Moreover, the ROM was parameterized by POD-greedy sampling, and a case study was performed where the kinetic data, i.e, the precursor decay constants and delayed-neutron fraction were assumed to be the model parameters of interest.
For the parameterized ROM, the maximum relative error in the assembly power was $1\times 10^{-3}$\%, which implies that the developed ROM can be reliably used to accelerate parametric and uncertainty quantification studies involving non-linear problems.
The computational gain from the MDEIM-ROM was compared with that of a direct projection ROM and showed that employing the MDEIM is computationally advantageous, and the advantage grows with dimension of the problem.
The method was implemented in the open-source deterministic transport code Detran\cite{roberts2014advanced} where only access to the system operator was needed to construct the ROM, and; hence, the method is trivially invasive and could easily be implemented in many modern deterministic diffusion codes.

@@

\begin{keywords}
Galerkin projection; POD; Non-linear reduced-order models; Neutronic Transient; Discrete Empirical Interpolation Method
\end{keywords}

\section{Introduction}
Modeling and simulation of reactor physics is crucial to understanding nuclear reactors' behavior under different conditions and scenarios and hence ensure reliable designs and well-established safety limits.
The mathematical models governing reactor physics are described by a set of equations that are challenging to solve analytically due to the geometric complexity of nuclear reactors except for simplified cases.
Alternatively, these systems are solved by employing numerical discretization that relies on discretizing the system along each independent variable and the solution is provided only on these discrete points.
To improve the method accuracy, the system needs to be solved over a very fine mesh.
Although this results in a more reliable solution, it leads to a growing space dimension and accordingly large computational cost.
This problem of the computational cost becomes inevitably prohibitive with applications that require repetitive execution of the simulation such as uncertainty quantification, and design optimization.
In such studies, the high-fidelity model has to be solved at multiple parameter points which could be very large.
Hence, surrogate models that are cheap, yet accurate, are sought to provide a rapid and reliable approximation of the solution of these systems.
The premise is that this lower-order model will capture the essential features of the system and preserve the input-output mapping as much as possible.

In the application of our interest here, neutronic transients, the mathematical models are based on the diffusion approximation and the Boltzmann transport equation.
For realistic modeling, these neutron transport models need to be coupled with different physics models to account for their feedback.
For example, thermal-hydraulics feedback plays an essential role in reactor operation.
Changes in the fuel temperature alter the cross sections due to Doppler broadening \cite{stacey2018nuclear}, which in turn impacts the neutron population and hence the fission rate and the rate of heat generation.
Multi-physics modeling accounts for these different feedback mechanisms; however, it poses a challenge in terms of the computational cost,
since it entails solving coupled systems for different fields such as neutron flux, temperature, and flow rate.
Moreover, the neutron transport (or diffusion) equation is no longer a linear one due to the indirect dependence of the cross section on the neutron flux.
Two approaches are used to solve such systems.
The most commonly used one is a serial coupling of two or more separate codes in such a way that the output of one at a given time step, such as the flux distribution, is used as an input to the other, e.g., the temperature (which can be used then to update the cross section and so on).
This procedure is commonly referred to as operator splitting or explicit coupling.
One drawback of operator splitting is that time integration requires small steps to avoid stability issues, thus leading to high computational time. 
Alternatively, tight coupling in time between the physics can be maintained by using an implicit scheme that solves a non-linear iteration in each time step, also increasing the computational time.
A modern algorithm such as Jacobian free Newton-Krylov (JFNK) \cite{gaston2009parallel} can be used to perform this implicit coupling.
Another popular approach is to use fixed-point iteration, with an implicit time scheme to iterate between difference physics within a time step until convergence is satisfied, which is computationally taxing.
The aim of this work is to use a reduced-order model to overcome the high computational expense of these non-linear models.
In particular, the ROM was developed using the POD-Galerkin projection.
The method belongs to the projection-based family and can be described as a physics-driven method since it requires knowledge of the original governing equation as will be shown in Sec.\ref{sec:POD}.
In a previous work \cite{elzohery2021modeling}, we applied the method to the time-dependent diffusion equation where there was no coupling with other models and hence the application was linear.
In this study, the developed ROM was extended to treat the non-linearity induced by multi-physics modeling.
To improve the computational efficiency of the developed ROM, a hyper-reduction technique known as the Discrete Empirical Interpolation Method (DEIM) was used in conjunction with the POD-Galerkin projection.
The primary purpose of using the DEIM is to obtain an offline-online decomposition of the ROM to avoid repetitive projection of the problem operator at each iteration within the time step.
The offline stage occurs only once and comprises the expensive step of the ROM.
On the other hand, the online stage involves solving a reduced number of equations at each time step with a computational cost independent of the FOM dimension and hence cheaper.
The LRA benchmark was used as a demonstrative application where the diffusion equation is coupled with a simplified thermal feedback model.

%The following discussion include a general overview of the methods and algorithms, then the specific application was described after which its results are presented.

@@

Although the problem tackled here does not include a non-linear term, the DEIM idea is exploited to approximate time-dependent operators to avoid computing the projected operator $\mathbf{A}_r(t)$ at each time step, which requires number of FLOPS dependent on $n$.
This is known as the matrix version of the DEIM, i.e, MDEIM and it was suggested in \cite{carlberg2012efficient, benner2015survey}.
To illustrate, consider the operator $\mathbf{A}(\boldsymbol{t}) \in \mathbb{R}^{n\times n}$.
The DEIM procedures can be used to approximate the operator by a summation of few time independent matrices, each weighted by a temporal coefficient.
First, the matrix can be cast in a vector form by stacking its elements in one vector $\mathbf{a}(t) \in \mathbb{R}^{n^2}$.
This vector can be treated in the same exact way as the function $\mathbf{f}(t)$ in the previous section, i.e, $\mathbf{a}(\mu) = \mathbf{\Psi}\boldsymbol{\theta}(t)$.
Snapshots of the serialized matrix are collected at different time instances and the reduced subspace $\mathbf{\Psi}$ and the interpolation indices are computed following algorithm \ref{alg:DEIM}.
Again, the corresponding coefficients are obtained by solving Eq.~\ref{eq:DEIM coefficients}. 
Next, each vector in $\mathbf{\Psi}$ can be mapped back to an $n \times n$ matrix, i.e, $\textbf{A}_i$.
Hence, the matrix is approximated as $\mathbf{\hat{A}}(t)=\sum_{i=0}^k \theta_i(t) \mathbf{A}_i$ and the projection term is computed as
\begin{equation}
	\mathbf{U}^T\mathbf{A}\mathbf{U} = \sum_{i=0}^k \mathbf{\theta}_i(t)\mathbf{U}^T\mathbf{A}_i\mathbf{U} \, .
	\label{Eq:deim matrix projection}
\end{equation}
As pointed out in Ref.\cite{negri2015efficient}, the Weyl\textendash Mirsky theorem \cite{weyl1912asymptotische} ensures that the singular values of the approximated matrix approach that of the original as $k$ increases 
\begin{equation}
	|\sigma_i - \hat{\sigma_i}  |\ \le \|\mathbf{A} - \hat{\mathbf{A}}\|_\text{F}
\end{equation}
where $\sigma_i$ and $\hat{\sigma_i}$ denote the $i$th singular value of $\mathbf{A}$ and $\hat{\mathbf{A}}$ respectively.
%Moreover, some properties of the original matrix such as positive definiteness can be preserved \cite{carlberg2015preserving}, if necessary, by imposing some constrains to the problem in Eq.~\ref{eq:DEIM coefficients}.

@@

%-----------------------------------------------------------------------------------------------------%
\section{Application}

\subsection{The time-dependent diffusion equation}
\label{sec:fom}
In this work, the time dependent diffusion equation was used as the mathematical model of neutronic transients in nuclear reactors, i.e FOM, and for which an approximate ROM is sought.
The multi-group version of the equation is represented as
\begin{equation}
	\begin{split}
	\frac{1}{v_g} \frac{\partial\phi_g}{\partial t} &= 
	\nabla. D_g\nabla\phi_g - \Sigma_{tg}\phi_g + \sum_{g^\prime=1}^{G}{\Sigma_{sg^\prime g}}\phi_{g^\prime} \\
	&+ \chi_g(1-\beta)\sum_{g^\prime = 1}^{G}\nu\Sigma_{fg^\prime}\phi_{g^\prime}  
	+ \chi_g\sum_{i=1}^{P} \lambda_i C_i \, ,
	\end{split}
	\label{time_diff}
\end{equation}
where 
\begin{equation*}
	\frac{\partial C_i}{\partial t} = -\lambda_i C_i + \beta_i \sum_{g^\prime=1}^{G} \nu\Sigma_{fg^\prime} \phi_{g^\prime} \, ,
\end{equation*}
and $\phi_g$, $D_g$, $\Sigma_{ag}$, $\nu\Sigma_{fg}$ and $v_g$ are the neutron flux, diffusion coefficient, the absorption cross section, the production cross section and the average neutron velocity in group g, respectively. $\Sigma_{sg^\prime g}$ is the scattering cross section from group $g^\prime$ to group $g$ 
and $\chi_g$ is the probability that a fission neutron will be born in group $g$.
$C_i$ is the $i$th precursor concentration, $\beta_i$ is the fraction of delayed neutrons in group $i$, $\lambda_i$ is the $i$th group decay constant.
%The system can be represented compactly in a matrix form as 
%\begin{equation}
%\frac{\partial{y}(r, t)}{\partial{t}} = \mathcal{A}{y}(r, t) \, ,
%\end{equation}
%where the stacked unknowns are
%\begin{equation*}
%	{y} = 
%	\begin{pmatrix}
%	\phi_1(r, t) \\
%	\vdots \\
%	\phi_G(r, t)  \\
%	{C}_1(r, t) \\
%	\vdots \\
%	{C}_I(r, t) \\
%	\end{pmatrix} \, .
%\end{equation*}

%\begin{equation*}
%\mathcal{A} = \left(\begin{array}{@{}c|c@{}}
%\begin{matrix}
%v_1(\nabla \cdot D_1\nabla +\Sigma_{11}) & v_1\Sigma_{12} & \hdots  &v_1\Sigma_{1G} \\
%v_2\Sigma_{21}        &    v_2(\nabla\cdot D_2\nabla +\Sigma_{22})            & \hdots  &v_2\Sigma_{2G} \\
%\vdots   &   \vdots   &  \ddots  &\vdots  \\
%v_G\Sigma_{G1}        &  v_G\Sigma_{G2} & \hdots &  v_G(\nabla \cdot D_G\nabla +\Sigma_{GG})     \\
%\end{matrix}
%&\begin{matrix}
%v_1 f_{11} & \hdots &   v_1 f_{1I} \\
%\vdots   & \ddots & \vdots \\
%v_G f_{G1} & \hdots &   v_G f_{GI}
%\end{matrix}
%\\
%\hline
%\begin{matrix}
%p_{11} &  &  & p_{12} & & & & \hdots  &   & p_{1G} \\
%\vdots &  &  & & \ddots & & & \vdots&   & \vdots \\
%p_{I1} &  &  & p_{I2} &  &   & & \hdots&  & p_{IG} 
%\end{matrix}
%&
%\begin{matrix}
%-\lambda_1 & \hdots & 0 \\
%\vdots & \ddots  &  \vdots \\
%0  &  \hdots  & -\lambda_I
%\end{matrix}
%\end{array}\right)
%\end{equation*}
%where 
%\begin{subequations}
%	\begin{equation*}
%	\Sigma{gg^\prime} = (1 -\beta)\chi_g\nu\Sigma{fg^\prime} + \Sigma_{s_{gg^\prime}} \, ,
%	\end{equation*}
%	\begin{equation*}
%	f_{ig^\prime} = \chi_g\lambda_i \, ,
%	\end{equation*}
%\end{subequations}
%\vspace{-2 pt}
%and
%\vspace{-2 pt}
%\begin{equation*}
%p_{ig} = \beta_i \nu\Sigma_{fg} \, .
%\end{equation*}
Upon spatial discretization, the resulting system of ODEs can be compactly represented in a matrix form as  
\begin{equation}
	\frac{d\mathbf{y}(t)}{dt} = \mathbf{A}\mathbf{y}(t) \, ,
	\label{eq:transient fom}
\end{equation}
where $\mathbf{y}\in\mathbb{R}^{N}$ is stacked vector of the flux and the precursors concentration, $\mathbf{A}\in\mathbb{R}^{N\times N}$ and $N = (I + G)\times n$ with $n$ is the number of spatial unknowns arising from discretization, $G$ is the number of energy groups, and $I$ is the number of precursors groups.
The operator $\mathbf{A}$ can be written in a block matrix form as
\begin{equation}
	\mathbf{A} =  \left(\begin{array}{@{}c|c@{}}
	\mathbf{L} & \mathbf{D} \\
	\hline
	\mathbf{F} & \mathbf{P}
	\end{array}\right) \, .
\end{equation} 
where $\mathbf{L} \in \mathbb{R}^{(n\times G)\times (n\times G)}$ represents the diffusion operator,
$\mathbf{D} \in \mathbb{R}^{(n\times G)\times (n\times I)}$ is a block diagonal matrix representing the delayed neutrons production operator,
$\mathbf{P} \in \mathbb{R}^{(n\times I)\times ((n\times G)}$ is a diagonal matrix represents the precursors concentration loss operator, 
and $\mathbf{F}\in \mathbb{R}^{(n\times I)\times ((n\times I)}$ represents the precursors production operator.
%Thus, the system in Eq.~\ref{eq:transient fom} is the full-order model and it can be solved using a suitable time-scheme.
%However, it is expensive to solve such systems for fine spatial and time grids, since a linear system of a size proportional to the spatial discretization has to be solved for each time step.
%Hence, a reduced-order model to approximate it is justified.

%--------------------------------------------------------------------------------------%
\subsection{Problem Description}

@@

\section{ROM}
\subsection{ROM Implementation}

The POD-Galerkin projection procedures described in Sec.~\ref{sec:POD} were applied to the system in Eq.~\ref{eq:transient fom} to approximate the group flux and the core power of the LRA benchmark with increasing the spatial fidelity, i.e, spatial grids of $22\times 22$, $44\times44$ and $55 \times 55$ were used.\\

@@

\subsection{ROM Results}

@@

\begin{figure}[H]
	\includegraphics[width=1.0\linewidth]{../figures/LRA_power_deim_relative_error.pdf}
	\caption{Relative error of the power predicted by MDEIM-ROM}
	\label{fig:deim power error}
\end{figure}
To understand the impact of employing the MDEIM on the error, the results are compared with these of a direct-projection ROM, i.e., with performing direct operator projection at each time step.
The relative difference of the power between both models is shown in Fig.~\ref{fig:rom-deim power error}.
The shown errors implies that MDEIM accurately approximates the problem operator.

@@

The MDEIM-ROM is sought to predict the power at every $0.001$ s during the transient, however, a time step of $0.005$ s was used in the greedy procedures.
Shown in Fig.~\ref{fig:lra greedy error} is the flux error convergence with the greedy iterations.
As shown, the maximum size of the POD space, i.e., 60, is reached before the error approaches the specified tolerance.
The resulting POD basis was stored and used for testing the ROM.

@@

\bibliography{interactnlmsample}

>>>>

\newif\ifDRAFT
\DRAFTtrue
%\DRAFTfalse

\ifDRAFT
\documentclass[review,number,sort&compress,12pt]{elsarticle}
%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}
\usepackage{lineno}
\modulolinenumbers[5]
\newcommand{\picwidth}{0.9\textwidth}
\linenumbers
\else 
\documentclass[final,5p,11pt,twocolumn]{elsarticle}
\newcommand{\picwidth}{0.48\textwidth}
\fi

\usepackage{algorithm,algorithmic}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{multirow}
\usepackage{booktabs,tabularx}
\usepackage{threeparttable}

\DeclareMathOperator{\argmaxH}{argmax} 
\DeclareMathOperator*{\argmin}{argmin}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\newcommand\abs[1]{\left|#1\right|}

%% `Elsevier LaTeX' style
\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%
\journal{Annals of Nuclear Energy}
\begin{document}
\setcitestyle{square}
%------------------------------------------------------------------------------
% 
%------------------------------------------------------------------------------
\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%

\title{Model-order reduction of non-linear time-dependent neutronics via POD-Galerkin projection and DEIM}
%% \tnotetext[label1]{}
\author[focal]{Rabab Elzohery\corref{cor}}
\ead{Rababelzohery@ksu.edu}
%\ead{http://www.rine.u-fukui.ac.jp}
\author{Jeremy Roberts}
%% \fntext[label2]{}
\cortext[cor1]{Corresponding author}
\address {The Alan Levin Department of Mechanical and Nuclear Engineering,
	Kansas State University \\
	3002 Rathbone Hall, Manhattan, KS 66506, US \\ rababelzohery@ksu.edu, jaroberts@ksu.edu}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

%------------------------------------------------------------------------------
% 
%------------------------------------------------------------------------------
\begin{abstract}
Presented here is an application of model-order reduction to non-linear neutronic transients.
The effort described extends previous work\cite{elzohery2021exploring} aimed to develop and assess a ROM framework that can be used for different purposes, such as design optimization and propagating input parameter uncertainties like nuclear data.
Here, the target applications is non-linear.
The ROM was built with the use of the POD-Galerkin projection method and the matrix version of the Discrete Empirical Interpolation Method (MDEIM) for improved efficiency.
The MDEIM is used to approximate the projection of the problem operator that otherwise must be computed at each non-linear iteration within each time step, which increases the ROM cost.
By applying the method to the LRA benchmark, speed ups from 6 to 40 were achieved depending on the full-order model (FOM) spatial fidelity.
The maximum relative error in the assemblies power was in the order of $1\times10^{-5}$\%.
Moreover, the ROM was parameterized by POD-greedy sampling, and a case study was performed where the kinetic data, i.e, the precursor decay constants and delayed-neutron fraction were assumed to be the model parameters of interest.
For the parameterized ROM, the maximum relative error in the assembly power was $1\times 10^{-3}$\%, 
The computational gain from the MDEIM-ROM was compared with that of a direct projection ROM and showed that employing the MDEIM is computationally advantageous, and the advantage grows with dimension of the problem.
The method was implemented in the open-source deterministic transport code Detran\cite{roberts2014advanced} where only access to the system operator was needed to construct the ROM, and; hence, the method is trivially invasive and could easily be implemented in many modern deterministic diffusion codes.

@@

%------------------------------------------------------------------------------
% 
%------------------------------------------------------------------------------
\begin{keyword}
	Reduced-Order Models; POD-Galerkin Projection; Discrete Empirical Interpolation Method; Non-linear Neutronic Transients
\end{keyword}

\end{frontmatter}

%%
%% Start line numbering here if you want
%%
%------------------------------------------------------------------------------
% 
%------------------------------------------------------------------------------
%\linenumbers

%------------------------------------------------------------------------------
%% main text
%------------------------------------------------------------------------------
\section{Introduction\label{sect::intro}}

Modeling and simulation of reactor physics is crucial to understanding nuclear reactors' behavior under different conditions and scenarios and hence ensure reliable designs and well-established safety limits.
The mathematical models governing reactor physics are described by a set of equations that are challenging to solve analytically due to the geometric complexity of nuclear reactors except for simplified cases.
Alternatively, these systems are solved by employing numerical discretization that relies on discretizing the system along each independent variable and the solution is provided only on these discrete points.
To improve the method accuracy, the system needs to be solved over a very fine mesh.
Although this results in a more reliable solution, it leads to a growing space dimension and accordingly large computational cost.
This problem of the computational cost becomes inevitably prohibitive with applications that require repetitive execution of the simulation such as uncertainty quantification, and design optimization.
In such studies, the high-fidelity model has to be solved at multiple parameter points which could be very large.
Hence, surrogate models that are cheap, yet accurate, are sought to provide a rapid and reliable approximation of the solution of these systems.
The premise is that this lower-order model will capture the essential features of the system and preserve the input-output mapping as much as possible.

In the application of our interest here, neutronic transients, the mathematical models are based on the diffusion approximation and the Boltzmann transport equation.
For realistic modeling, these neutron transport models need to be coupled with different physics models to account for their feedback.
For example, thermal-hydraulics feedback plays an essential role in reactor operation.
Changes in the fuel temperature alter the cross sections due to Doppler broadening \cite{stacey2018nuclear}, which in turn impacts the neutron population and hence the fission rate and the rate of heat generation.
Multi-physics modeling accounts for these different feedback mechanisms; however, it poses a challenge in terms of the computational cost,
since it entails solving coupled systems for different fields such as neutron flux, temperature, and flow rate.
Moreover, the neutron transport (or diffusion) equation is no longer a linear one due to the indirect dependence of the cross section on the neutron flux.
several approaches are used to solve such systems.
The most commonly used one is a serial coupling of two or more separate codes in such a way that the output of one at a given time step, such as the flux distribution, is used as an input to the other, e.g., the temperature (which can be used then to update the cross section and so on).
This procedure is commonly referred to as operator splitting or explicit coupling.
One drawback of operator splitting is that time integration requires small steps to avoid stability issues, thus leading to high computational time. 
Alternatively, tight coupling in time between the physics can be maintained by using an implicit scheme that solves a non-linear iteration in each time step, also increasing the computational time.
A modern algorithm such as Jacobian free Newton-Krylov (JFNK) \cite{gaston2009parallel} can be used to perform this implicit coupling.
Another popular approach is to use fixed-point iteration, with an implicit time scheme to iterate between difference physics within a time step until convergence is satisfied, which is computationally taxing.
The aim of this work is to use a reduced-order model to overcome the high computational expense of these non-linear models.
In particular, the ROM was developed using the POD-Galerkin projection.
The method belongs to the projection-based family and can be described as a physics-driven method since it requires knowledge of the original governing equation as will be shown in Sec.\ref{sec:POD}.
In a previous work \cite{elzohery2021exploring}, we applied the method to the time-dependent diffusion equation where there was no coupling with other models and hence the application was linear.
In this study, the developed ROM was extended to treat the non-linearity induced by multi-physics modeling.
To improve the computational efficiency of the developed ROM, a hyper-reduction technique known as the Discrete Empirical Interpolation Method (DEIM) was used in conjunction with the POD-Galerkin projection.
The primary purpose of using the DEIM is to obtain an offline-online decomposition of the ROM to avoid repetitive projection of the problem operator at each iteration within the time step.
The offline stage occurs only once and comprises the expensive step of the ROM.
On the other hand, the online stage involves solving a reduced number of equations at each time step with a computational cost independent of the FOM dimension and hence cheaper.
The LRA benchmark was used as a demonstrative application where the diffusion equation is coupled with a simplified thermal feedback model.

@@

Although the problem tackled here does not include a non-linear term, the DEIM idea is exploited to approximate time-dependent operators to avoid computing the projected operator $\mathbf{A}_r(t)$ at each time step, which requires number of FLOPS dependent on $n$.
This is known as the matrix version of the DEIM, i.e, MDEIM and it was suggested in \cite{carlberg2012efficient},\cite{benner2015survey}.
To illustrate, consider the operator $\mathbf{A}(\boldsymbol{t}) \in \mathbb{R}^{n\times n}$.
The DEIM procedures can be used to approximate the operator by a summation of few time independent matrices, each weighted by a temporal coefficient.
First, the matrix can be cast in a vector form by stacking its elements in one vector $\mathbf{a}(t) \in \mathbb{R}^{n^2}$.
This vector can be treated in the same exact way as the function $\mathbf{f}(t)$ in the previous section, i.e, $\mathbf{a}(\mu) = \mathbf{\Psi}\boldsymbol{\theta}(t)$.
Snapshots of the serialized matrix are collected at different time instances and the reduced subspace $\mathbf{\Psi}$ and the interpolation indices are computed following algorithm \ref{alg:DEIM}.
Again, the corresponding coefficients are obtained by solving Eq.~\ref{eq:DEIM coefficients}. 
Next, each vector in $\mathbf{\Psi}$ can be mapped back to an $n \times n$ matrix, i.e, $\textbf{A}_i$.
Hence, the matrix is approximated as $\mathbf{\hat{A}}(t)=\sum_{i=0}^k \theta_i(t) \mathbf{A}_i$ and the projection term is computed as
\begin{equation}
\mathbf{U}^T\mathbf{A}\mathbf{U} = \sum_{i=0}^k \mathbf{\theta}_i(t)\mathbf{U}^T\mathbf{A}_i\mathbf{U} \, .
\label{Eq:deim matrix projection}
\end{equation}
As pointed out in Ref.\cite{negri2015efficient}, the Weyl\textendash Mirsky theorem \cite{weyl1912asymptotische} ensures that the singular values of the approximated matrix approach that of the original as $k$ increases 
\begin{equation}
|\sigma_i - \hat{\sigma_i}  |\ \le \|\mathbf{A} - \hat{\mathbf{A}}\|_\text{F}
\end{equation}
where $\sigma_i$ and $\hat{\sigma_i}$ denote the $i$th singular value of $\mathbf{A}$ and $\hat{\mathbf{A}}$ respectively.
%Moreover, some properties of the original matrix such as positive definiteness can be preserved \cite{carlberg2015preserving}, if necessary, by imposing some constrains to the problem in Eq.~\ref{eq:DEIM coefficients}.

@@

%-----------------------------------------------------------------------------------------------------%
\section{The time-dependent diffusion equation}
\label{sec:fom}
In this work, the time dependent diffusion equation was used as the mathematical model of neutronic transients in nuclear reactors, i.e FOM, and for which an approximate ROM is sought.
The multi-group version of the equation is represented as
\begin{equation}
\begin{split}
\frac{1}{v_g} \frac{\partial\phi_g}{\partial t} &= 
\nabla. D_g\nabla\phi_g - \Sigma_{tg}\phi_g + \sum_{g^\prime=1}^{G}{\Sigma_{sg^\prime g}}\phi_{g^\prime} \\
&+ \chi_g(1-\beta)\sum_{g^\prime = 1}^{G}\nu\Sigma_{fg^\prime}\phi_{g^\prime}  
+ \chi_g\sum_{i=1}^{P} \lambda_i C_i \, ,
\end{split}
\label{time_diff}
\end{equation}
where 
\begin{equation*}
\frac{\partial C_i}{\partial t} = -\lambda_i C_i + \beta_i \sum_{g^\prime=1}^{G} \nu\Sigma_{fg^\prime} \phi_{g^\prime} \, ,
\end{equation*}
and $\phi_g$, $D_g$, $\Sigma_{ag}$, $\nu\Sigma_{fg}$ and $v_g$ are the neutron flux, diffusion coefficient, the absorption cross section, the production cross section and the average neutron velocity in group g, respectively. $\Sigma_{sg^\prime g}$ is the scattering cross section from group $g^\prime$ to group $g$ 
and $\chi_g$ is the probability that a fission neutron will be born in group $g$.
$C_i$ is the $i$th precursor concentration, $\beta_i$ is the fraction of delayed neutrons in group $i$, $\lambda_i$ is the $i$th group decay constant.
Upon spatial discretization, the resulting system of ODEs can be compactly represented in a matrix form as  
\begin{equation}
\frac{d\mathbf{y}(t)}{dt} = \mathbf{A}\mathbf{y}(t) \, ,
\label{eq:transient fom}
\end{equation}
where $\mathbf{y}\in\mathbb{R}^{N}$ is stacked vector of the flux and the precursors concentration, $\mathbf{A}\in\mathbb{R}^{N\times N}$ and $N = (I + G)\times n$ with $n$ is the number of spatial unknowns arising from discretization, $G$ is the number of energy groups, and $I$ is the number of precursors groups.
The operator $\mathbf{A}$ can be written in a block matrix form as
\begin{equation}
\mathbf{A} =  \left(\begin{array}{@{}c|c@{}}
\mathbf{L} & \mathbf{D} \\
\hline
\mathbf{F} & \mathbf{P}
\end{array}\right) \, .
\end{equation} 
where $\mathbf{L} \in \mathbb{R}^{(n\times G)\times (n\times G)}$ represents the diffusion operator,
$\mathbf{D} \in \mathbb{R}^{(n\times G)\times (n\times I)}$ is a block diagonal matrix representing the delayed neutrons production operator,
$\mathbf{P} \in \mathbb{R}^{(n\times I)\times ((n\times G)}$ is a diagonal matrix represents the precursors concentration loss operator, 
and $\mathbf{F}\in \mathbb{R}^{(n\times I)\times ((n\times I)}$ represents the precursors production operator.
%Thus, the system in Eq.~\ref{eq:transient fom} is the full-order model and it can be solved using a suitable time-scheme.
%However, it is expensive to solve such systems for fine spatial and time grids, since a linear system of a size proportional to the spatial discretization has to be solved for each time step.
%Hence, a reduced-order model to approximate it is justified.

%--------------------------------------------------------------------------------------%
\section{Application}
\subsection{Problem Description}

@@

\subsection{ROM Implementation}

The POD-Galerkin projection procedures described in Sec.~\ref{sec:POD} were applied to the system in Eq.~\ref{eq:transient fom} to approximate the group flux and the core power of the LRA benchmark with increasing the spatial fidelity, i.e, spatial grids of $22\times 22$, $44\times44$, and $55 \times 55$ were used.\\

@@

\subsubsection{ROM Results}

@@

\begin{figure}[H]
	\includegraphics[width=1.0\linewidth]{../figures/_.pdf}
	\caption{Relative error of the power predicted by MDEIM-ROM}
	\label{fig:deim power error}
\end{figure}
To understand the impact of employing the MDEIM on the error, the results are compared with these of a direct-projection ROM, i.e., with performing direct operator projection at each time step.
The relative difference of the power between both models is shown in Fig.~\ref{fig:rom-deim power error}.
The shown errors implies that MDEIM accurately approximates the problem operator.

@@

The MDEIM-ROM is sought to predict the power at every $0.001$ s during the transient.
However, a time step of $0.005$ s was used in the greedy procedures, since, as shown in \cite{elzohery2021exploring}, using a smaller time step in the generation of the greedy POD basis leads to faster error convergence.
Shown in Fig.~\ref{fig:lra greedy error} is the flux error convergence with the greedy iterations.
As shown, the maximum size of the POD space, i.e., 60, is reached before the error approaches the specified tolerance.
The resulting POD basis was stored and used for testing the ROM.

@@

\section*{References}

\bibliography{mybibfile}
